---
title: "Chromatin accessibility vs 5mC analysis - using Cochran-Mantel-Haenszel test"
author: "Elisa Kreibich"
date: "23/03/2022"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: hide
# editor_options:
#   chunk_output_type: inline
  # html_document:
  #   toc: true
  #   toc_float: true
  #   code_folding: hide
editor_options: 
  chunk_output_type: inline
---
**Disclaimer about nomenclature**  
In the manuscript, we call the different states "sites with xyz 5mC-CA-association". Here, we call them in their original nomenclature:  

* antagonist = negative 5mC-CA-association  
* neutral = no 5mC-CA-association  
* agonist = positive 5mC-CA-association  


# Introduction
### Aim
Analysis of SMF WT ES data to identify antagonist and neutral sites, define them and analyse their behavior in DNMT and TET TKOs.  

### Input
**Files needed to run this script and which scripts to run to create those:**

| Input file name | Information | Script to create |
|:----------------|:------------|:-----------------|
| Final_data_tibble_CA_***DATE*** _ES_NO_TKO_DE_TETTKO_NO_cObin ***BINCOVERAGECUTOFF***_cOCMH***CMHCOVERAGECUTOFF***_states.rds  | Final tibble with states and calculated changes | Make_Final_CG_CMH_data_tibbles_ES_plus_TKOs.Rmd  |  

\

### Output
Plots of the following analyses:  

* Quick checks on coverage, 5mC distribution etc in WT and TKOs
* Changes in accessibility in DNMT TKOs
* Changes in accessibility in TET TKOs 

\

### State definitions
**States are defined as followed:**  

* **state** 
    * antagonist: COR <= 0.5 | pvalue < 0.05 | ICR == FALSE 
    * neutral: COR > 0.5 | no pvalue cutoff | ICR == FALSE 
    * agonist: COR >= 2 | pvalue < 0.05 | ICR == FALSE
    * ICR: ICR == TRUE

* **state2**
    * antagonist: COR <= 0.5 | pvalue < 0.05 | ICR == FALSE  
    * neutral+: COR > 0.9 & < 1.1 | no pvalue cutoff | ICR == FALSE 
    * neutral: COR > 0.5 | no pvalue cutoff | ICR == FALSE 
    * agonist: COR >= 2 | pvalue < 0.05 | ICR == FALSE
    * ICR: ICR == TRUE

* **state0**
    * antagonist: COR <= 0.5 | pvalue < 0.05
    * neutral: COR > 0.5 | no pvalue cutoff  
    * agonist: COR >= 2 | pvalue < 0.05 

\

### Additional information on this version:  

* 101 bp collection window around CpG  
* 10x coverage cutoff per bin for NWCGW and DGCHN methylation  
* 30x coverage cutoff for SM analysis 
* bins most be covered by at least 2 replicates  
* average methylation (\*\_wmean_me\_\*) == weighted mean (wmean) of all replicates  
* average accessibility (\*\_wmean_acc\_\*) == mean of GpC methylation in bin + weighted mean (wmean) of all replicates  

\

# Data Analysis
## Set environment
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
```{r}
WD = '/g/krebs/kreibich/Kreibich_2023_5mC_at_enhancers/'
```

### Load libraries
```{r load libraries, include=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(patchwork))

source(paste0(WD, 'scripts/utilities/COLORS.R'))
#Load arguments
source(paste0(WD, 'scripts/utilities/Input_arguments_CA.R'))
```

### Set arguments
```{r arguments}
INDIR           <- 'data_results/'
OUTDIR          <- 'data_results/'
dir.create(OUTDIR)

DATE            <- Sys.Date()

#Arguments for WT input data set
# SOI           <- c("SMF_MM_ES_NO_R1", "SMF_MM_ES_NO_R2", "SMF_MM_ES_NO_R5a6")
# SOI_REPS      <- str_extract(SOI[-1], pattern = "R.{1,3}")
# INPUT_DATE_WT <- c("2022-10-27")  #Date of Final WT tibble creation

#Arguments for Final WT+TKOs input data set
SAMPLE_OI       <- "ES"
SAMPLE_OI_full  <- c("ES_NO", "TKO_DE", "TETTKO_NO")
SAMPLES_OI_2    <- c("ES WT", "DNMT TKO", "TET TKO")
INPUT_DATE      <- c("2022-10-24")  #Date of Final WT+TKO tibble creation

```

### Load data
#### Final data tibble WT and TKOs
```{r}
files   <- list.files(path = paste0(WD, INDIR),
                    pattern =  paste('Final_data_tibble_CA', INPUT_DATE, 
                         ".{1,}",
                         paste0("cObin", cO_bin),
                         paste0("cOCMH", cO_CMH),
                         ".{1,}.rds",
                         sep = '_'),
                    full.names = TRUE)
files   <- files[grepl(paste(SAMPLE_OI_full, collapse = "_"), files)]
data    <- readRDS(files[grepl("_states.rds", files)])
# data_gr  <- readRDS(files[grepl("_states_gr.rds", files)])
```
**Tibble mutations - with states**
```{r}
# Filter out high and low methylated regions
data <- data %>%
  filter(!(grepl("_5mC", state)))

# Filter out Inactive and CTCF annotations to focus on active regulatory regions
data_f <- data %>%
  filter(!(chromHMMcl %in% c("C1_CTCF", "C2_Inactive")))
```

\

## #1 Quick check - WT + TKOs
### Coverage at bins - all cell lines
```{r message=FALSE, warning=FALSE}
vars1 <- names(data)[grepl("mean_cov", names(data))]
vars1_n <- str_remove(vars1[1], "^\\w{1,}_(NO|DE)_")

p_u <- data %>% 
  select(bin, vars1) %>% 
  pivot_longer(cols = -bin,
               names_to = "cell_line",
               values_to = vars1_n) %>% 
  mutate(cell_line = str_remove(cell_line, paste0("_", vars1_n))) %>% 
  mutate(cell_line = case_when(
    cell_line == "ES_NO" ~ SAMPLES_OI_2[1],
    cell_line == "TKO_DE" ~ SAMPLES_OI_2[2],
    cell_line == "TKO_NO" ~ SAMPLES_OI_2[2],
    cell_line == "TETTKO_NO" ~ SAMPLES_OI_2[3]
  )) %>% 
  mutate(cell_line = factor(cell_line, levels = SAMPLES_OI_2)) %>% 
  ggplot(aes(cell_line, get(vars1_n), fill = cell_line)) +
  geom_violin() +
  geom_boxplot(fill = "white",  width = .1, alpha = 0.8, outlier.shape = NA, coef = 0) +
  scale_y_log10() +
  scale_fill_manual(values = COLORS_TYPES) +
  theme_EK() +
  labs(y = vars1_n,
       subtitle = "Final data")
p_u

```

\

## #2 Analysis - WT & TKOs - Global 5mC Pertubations
Here I will do only the analyses for the agonist and neutral sites and will remove the agonist sites. 
The agonist sites are analysed in a separate script.

```{r}
data_f2 <- data_f %>% 
  filter(state != "agonist")
```

**Functions**
```{r}
ggplot.acc <- function(tbl = data_f2, CL1, CL2, CO_5mC, filter2){
  var1 <- c(
            names(tbl)[grepl("acc_cov_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))],
            names(tbl)[grepl("wmean_acc_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))],
            names(tbl)[grepl("wmean_me_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))]
            )
  var2 <- c(
            names(tbl)[grepl("acc_cov_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))],
            names(tbl)[grepl("wmean_acc_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))],
            names(tbl)[grepl("wmean_me_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))]
            )   
  
  data_tbl <- if(CL2 == "TKO"){
    tbl %>% 
      mutate(NWCGW_change = get(var1[3]))
  } else if(CL2 == "TETTKO"){
     tbl %>% 
      mutate(NWCGW_change = get(var2[3]) - get(var1[3]))
  } 
  data_tbl <- data_tbl %>% 
    filter(NWCGW_change >= CO_5mC) %>% 
    filter(state2 != "neutral") %>%
    filter(state2 != filter2)
  
  count <- data_tbl %>% 
    count(state2) %>% 
    mutate(label = paste0(state2, " (", n, ")")) %>% 
    pull(label)
  
  data_tbl %>% 
    ggplot(aes(x = state2, y = -(get(var2[2]) - get(var1[2]))*100, fill = state2)) +
    geom_violin() +
    geom_boxplot(fill = "white", color = "black", width = 0.2, alpha = 0.8, outlier.alpha = 0, coef = 0) +
    stat_compare_means(label = "p.signif", size = 3, ref.group = "neutral+") +
    stat_compare_means(ref.group = "neutral+", label.y = -Inf) +
    scale_fill_manual(values = COLORS_STATE_v3s) +
    scale_color_gradient(low = "grey75", high = "grey5") +
    labs(subtitle = paste("SMF (1-accessibility) change in", CL1, "vs", CL2),
         y = paste0("delta SMF (1-accessibility) (", CL2," - ", CL1, ") [%]"),
         x = "",
         caption = paste(count, collapse = "; ")) +
    theme_EK()
}
```

### TKO 
#### Global change in accessibility
```{r}
CL1       <- "ES"     #cell line ref
CL2       <- "TKO"    #cell line 2
CO_5mC    <- 0.3      #cutoff 5mC difference --> at least 30% methylation difference
TITLE     <- paste("delta accessibility", CL1, "vs", CL2)

filter2   <- ""   #filter out additional state 
pVio_TKO2 <- ggplot.acc(data, CL1, CL2, CO_5mC, filter2) + 
  ylim(-40,20) + 
  scale_fill_manual(values = COLORS_STATE_v3s[-1]) +
  ggtitle("no filters")
pVio_TKO2
# ggsave(plot = pVio_TKO2, filename = paste0(WD, OUTDIR, paste("AccChange", "TKO", "NOTfiltered", "wICR", "wAgonist", sep="_"), "_v2_", DATE, ".pdf"), width = 4.5, height = 5)

filter2   <- ""   #filter out additional state 
pVio_TKO1 <- ggplot.acc(data_f, CL1, CL2, CO_5mC, filter2) + 
  ylim(-40,20) + 
  scale_fill_manual(values = COLORS_STATE_v3s[-1]) +
  ggtitle("filtered for chromHMM")
pVio_TKO1
ggsave(plot = pVio_TKO1, filename = paste0(WD, OUTDIR, paste("AccChange", "TKO", "filtered_chromHMM", "wICR", "wAgonist", sep="_"), "_v", DATE, ".pdf"), width = 4.5, height = 5)

filter2   <- "agonist"   #filter out additional state 
pVio_TKO3 <- ggplot.acc(data_f2, CL1, CL2, CO_5mC, filter2) + 
  ylim(-40,20) + 
  scale_fill_manual(values = COLORS_STATE_v3s[-c(1,5)]) +
  ggtitle("filtered for chromHMM")
pVio_TKO3
ggsave(plot = pVio_TKO3, filename = paste0(WD, OUTDIR, paste("AccChange", "TKO", "filtered_chromHMM", "wICR", sep="_"), "_v2_", DATE, ".pdf"), width = 4.5, height = 5)

```

\

##### PValues of statistical test
```{r}
CL1       <- "ES"     #cell line ref
CL2       <- "TKO"    #cell line 2
CO_5mC    <- 0.3      #cutoff 5mC difference

df <- data_f %>% 
  mutate(NWCGW_change = ES_NO_wmean_me_bin,
         DGCHN_change = -(TKO_DE_wmean_acc_bin - ES_NO_wmean_acc_bin)) %>%
  filter(NWCGW_change >= CO_5mC) %>% 
  filter(state2 != "neutral")

compare_means(DGCHN_change ~ state2, df, method = "wilcox.test", ref.group = "neutral+")
# compare_means(DGCHN_change ~ state2, df, method = "t.test", ref.group = "neutral+")
```

\

#### Correlation change 5mC and accessibility
**Set-up data**
```{r fig.height=4, fig.width=12}
CL1     <- "ES"     #cell line ref
CL2     <- "TKO" #cell line 2
CO_5mC  <- 0.3   #cutoff 5mC difference
covCO   <- 10     #cutoff coverage

tbl <- data_f
  var1 <- c(
            names(tbl)[grepl("acc_cov_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))],
            names(tbl)[grepl("wmean_acc_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))],
            names(tbl)[grepl("wmean_me_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))]
            )
  var2 <- c(
            names(tbl)[grepl("acc_cov_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))],
            names(tbl)[grepl("wmean_acc_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))],
            names(tbl)[grepl("wmean_me_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))]
            )   
  
  data_tbl <- if(CL2 == "TKO"){
    tbl %>% 
      mutate(NWCGW_change = get(var1[3]))
  } else if(CL2 == "TETTKO"){
     tbl %>% 
      mutate(NWCGW_change = get(var2[3]) - get(var1[3]))
  } 
```

\

**Correlation score**
```{r}
data_tbl2 <- data_tbl %>% 
  filter(state2 != "neutral") %>%
  filter(state2 != "ICR") %>%
  filter(NWCGW_change >= 0) %>% 
  mutate(accCh = -(get(var2[2]) - get(var1[2])),
         log2FC = -(log2((get(var2[2])+0.001)/(get(var1[2])+0.001))))

cor_val <- data_tbl2 %>% 
  select(bin, state2, accCh, log2FC, NWCGW_change) %>% 
  distinct(bin, .keep_all = T) %>% 
  pivot_wider(names_from = "state2", values_from = c("accCh", "log2FC", "NWCGW_change"), names_sep = "_") %>% 
  select(-bin) %>% 
  corrr::correlate()
```

\

**Delta accessibility**
```{r}
# Both states
numbers <- data_tbl2 %>% 
  count(state2) %>% 
  mutate(label = paste(state2, "=", n)) 
  

p_AccCh <- data_tbl2 %>% 
    ggplot(aes(x = NWCGW_change*100, y = accCh*100, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.6, size = 2, show.legend = T) +
    geom_smooth(color = "black", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v3s[c(2,3,5)]) + 
    scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("SMF (1 - accessibility) change in", CL1, "vs", CL2),
         y = paste0("delta SMF (1 - CA) (", CL2," - ", CL1, ") [%]"),
         x = paste0("delta 5mC (", CL2," - ", CL1, ") [%]"),
         caption = paste("R neutral =", round(cor_val$`accCh_neutral+`[7],3),
                         "\nR antagonist =", round(cor_val$`accCh_antagonist`[8],3),
                         "\nR agonist =", round(cor_val$`accCh_agonist`[9],3),
                         "\n", paste(numbers$label, collapse = " | "))) +
    theme_EK() +
    facet_wrap(~state2, ncol = 3) 
p_AccCh

FILENAME = paste("AccChange", CL2, "lm", sep="_")
ADD = "_v1_"
# ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".png"), height = 4, width = 8)
# ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".pdf"), height = 4, width = 8)
```
```{r}
# Only antagonist
numbers <- data_tbl2 %>% 
  filter(state2 == "antagonist") %>% 
  nrow()
  
p_AccCh <- data_tbl2 %>% 
    filter(state2 == "antagonist") %>% 
    ggplot(aes(x = NWCGW_change*100, y = accCh*100, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = T) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v2[2:3]) + 
    scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("delta SMF (1-CA) (", CL2," - ", CL1, ") [%]"),
         x = paste0("delta 5mC (", CL1," - ", CL2, ") [%]"),
         caption = paste("R antagonist =", round(cor_val$`accCh_antagonist`[8],3), "| n =", numbers)) +
    theme_EK() +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

FILENAME = paste("AccChange", CL2, "lm", "antagonist", sep="_")
ADD = "_v1_"
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".png"), height = 4, width = 6.5)
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".pdf"), height = 4, width = 6.5)
```
```{r}
# Only agonist
numbers <- data_tbl2 %>% 
  filter(state2 == "agonist") %>% 
  nrow()
  
p_AccCh <- data_tbl2 %>% 
    filter(state2 == "agonist") %>% 
    ggplot(aes(x = NWCGW_change*100, y = accCh*100, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = T) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v3[5]) + 
    scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("delta accessibility (", CL2," - ", CL1, ") [%]"),
         x = paste0("delta 5mC (", CL1," - ", CL2, ") [%]"),
         caption = paste("R agonist =", round(cor_val$`accCh_agonist`[9],3), "| n =", numbers)) +
    theme_EK() +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

FILENAME = paste("AccChange", CL2, "lm", "agonist", sep="_")
ADD = "_v1_"
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".png"), height = 4, width = 6.5)
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".pdf"), height = 4, width = 6.5)
```
```{r}
# Only neutral+
numbers <- data_tbl2 %>% 
  filter(state2 == "neutral+") %>% 
  nrow()
  
p_AccCh <- data_tbl2 %>% 
    filter(state2 == "neutral+") %>% 
    ggplot(aes(x = NWCGW_change*100, y = accCh*100, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = T) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v3[2:3]) + 
    scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("delta SMF (1-CA) (", CL2," - ", CL1, ") [%]"),
         x = paste0("delta 5mC (", CL1," - ", CL2, ") [%]"),
         caption = paste("R neutral =", round(cor_val$`accCh_neutral+`[7],3), "| n =", numbers)) +
    theme_EK() +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

FILENAME = paste("AccChange", CL2, "lm", "neutral", sep="_")
ADD = "_v1_"
# ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".png"), height = 4, width = 6.5)
# ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".pdf"), height = 4, width = 6.5)
```

\

**log2(FC) accessibility**
```{r}
p_AccCh <- data_tbl2 %>% 
    ggplot(aes(x = NWCGW_change*100, y = log2FC, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = F) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v3[2:5]) + 
    # scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("log2 (fold change in accessibility) (", CL2," - ", CL1, ")"),
         x = paste0("delta 5mC (", CL2," - ", CL1, ") [%]"),
         caption = paste("R neutral =", round(cor_val$`log2FC_neutral+`[7],3),
                         "\nR antagonist =", round(cor_val$log2FC_antagonist[8],3),
                         "\nR agonist =", round(cor_val$log2FC_agonist[9],3))) +
    theme_EK() +
    # theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

```

\

##### Linear model
```{r}
df_A <- data_f2 %>% 
  mutate(NWCGW_change = ES_NO_wmean_me_bin,
         DGCHN_change = -(TKO_DE_wmean_acc_bin - ES_NO_wmean_acc_bin)) %>%
  filter(state2 == "antagonist")

TKO_A.lm <- lm(DGCHN_change ~ NWCGW_change, data = df_A)
summary(TKO_A.lm)

df_N <- data_f2 %>% 
  mutate(NWCGW_change = ES_NO_wmean_me_bin,
         DGCHN_change = -(TKO_DE_wmean_acc_bin - ES_NO_wmean_acc_bin)) %>%
  filter(state2 == "neutral+")

TKO_N.lm <- lm(DGCHN_change ~ NWCGW_change, data = df_N)
summary(TKO_N.lm)
```

\

### TETTKO
#### Global change in accessibility
```{r}
CL1       <- "ES"         #cell line ref
CL2       <- "TETTKO"    #cell line 2
CO_5mC    <- 0.3          #cutoff 5mC difference
FILENAME  <- paste("AccChange", "TET", sep="_")


filter2   <- ""   #filter out additional state 
ADD       <- paste("NOT_filtered", "wICR", "wAgonist", sep="_")
pVio_TET5 <- ggplot.acc(data, CL1, CL2, CO_5mC, filter2) + 
  ylim(-30,40) + 
  scale_fill_manual(values = COLORS_STATE_v3s[-1]) +
  ggtitle("no filters")
pVio_TET5
# ggsave(plot = pVio_TET5, filename = paste0(WD, OUTDIR, FILENAME, ADD, "_v", DATE, ".pdf"), width = 4.5, height = 5)


filter2   <- ""   #filter out additional state 
ADD       <- paste("filtered_chromHMM", "wICR", "wAgonist", sep="_")
pVio_TET1 <- ggplot.acc(data_f, CL1, CL2, CO_5mC, filter2) + 
  ylim(-30,40) + 
  scale_fill_manual(values = COLORS_STATE_v3s[-1]) +
  ggtitle("filtered for chromHMM")
pVio_TET1
ggsave(plot = pVio_TET1, filename = paste0(WD, OUTDIR, FILENAME, ADD, "_v", DATE, ".pdf"), width = 4.5, height = 5)

filter2   <- ""   #filter out additional state 
ADD       <- paste("filtered_chromHMM", "wICR", sep="_")
pVio_TET1 <- ggplot.acc(data_f2, CL1, CL2, CO_5mC, filter2) + 
  ylim(-30,40) + 
  scale_fill_manual(values = COLORS_STATE_v1s[-1]) +
  ggtitle("filtered for chromHMM")
pVio_TET1
ggsave(plot = pVio_TET1, filename = paste0(WD, OUTDIR, FILENAME, ADD, "_v", DATE, ".pdf"), width = 4.5, height = 5)
```

##### PValues of statistical test
```{r}
CL1     <- "ES"     #cell line ref
CL2     <- "TETTKO" #cell line 2
CO_5mC  <- 0.3      #cutoff 5mC difference

df <- data_f2 %>% 
  mutate(NWCGW_change = TETTKO_NO_wmean_me_bin - ES_NO_wmean_me_bin,
         DGCHN_change = 1-(TETTKO_NO_wmean_acc_bin - ES_NO_wmean_acc_bin)) %>%
  filter(NWCGW_change >= CO_5mC) %>% 
  filter(state2 != "neutral")

compare_means(DGCHN_change ~ state2, df, method = "wilcox.test", ref.group = "neutral+")
```

#### Correlation change 5mC and accessibility
**Set-up data**
```{r fig.height=4, fig.width=12}
CL1     <- "ES"     #cell line ref
CL2     <- "TETTKO" #cell line 2
CO_5mC  <- 0.3   #cutoff 5mC difference

tbl <- data_f
  var1 <- c(
            names(tbl)[grepl("acc_cov_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))],
            names(tbl)[grepl("wmean_acc_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))],
            names(tbl)[grepl("wmean_me_bin", names(tbl)) & grepl(paste0("^", CL1), names(tbl))]
            )
  var2 <- c(
            names(tbl)[grepl("acc_cov_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))],
            names(tbl)[grepl("wmean_acc_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))],
            names(tbl)[grepl("wmean_me_bin", names(tbl)) & grepl(paste0("^", CL2), names(tbl))]
            )   
  
  data_tbl <- if(CL2 == "TKO"){
    tbl %>% 
      mutate(NWCGW_change = get(var1[3]))
  } else if(CL2 == "TETTKO"){
     tbl %>% 
      mutate(NWCGW_change = get(var2[3]) - get(var1[3]))
  } 
```

\

**Correlation score**
```{r}
data_tbl2 <- data_tbl %>% 
  filter(state2 != "neutral") %>%
  filter(state2 != "ICR") %>%
  filter(NWCGW_change >= 0) %>% 
  mutate(accCh =  -(get(var2[2]) - get(var1[2])),
         log2FC = -(log2((get(var2[2])+0.001)/(get(var1[2])+0.001))))

cor_val <- data_tbl2 %>% 
  select(bin, state2, accCh, log2FC, NWCGW_change) %>% 
  distinct(bin, .keep_all = T) %>% 
  pivot_wider(names_from = "state2", values_from = c("accCh", "log2FC", "NWCGW_change"), names_sep = "_") %>% 
  select(-bin) %>% 
  corrr::correlate()

cor_val
```

\

**Delta accessibility**
```{r}
# Both states
numbers <- data_tbl2 %>% 
  count(state2) %>% 
  mutate(label = paste(state2, "=", n)) 
  
p_AccCh <- data_tbl2 %>% 
    ggplot(aes(x = NWCGW_change*100, y = accCh*100, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = T) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v3[2:5]) + 
    scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("delta SMF (1- CA) (", CL2," - ", CL1, ") [%]"),
         x = paste0("delta 5mC (", CL1," - ", CL2, ") [%]"),
         caption = paste("R neutral =", round(cor_val$`accCh_neutral+`[7],3),
                         "\nR antagonist =", round(cor_val$`accCh_antagonist`[8],3),
                         "\nR agonist =", round(cor_val$`accCh_agonist`[9],3),
                         "\n", paste(numbers$label, collapse = " | "))) +
    theme_EK() +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

FILENAME = paste("AccChange", CL2, "lm", sep="_")
ADD = "_v1_"
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".png"), height = 4, width = 7)
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".pdf"), height = 4, width = 7)
```
```{r}
# Only antagonist
numbers <- data_tbl2 %>% 
  filter(state2 == "antagonist") %>% 
  count(state2)
  
p_AccCh <- data_tbl2 %>% 
    filter(state2 == "antagonist") %>% 
    ggplot(aes(x = NWCGW_change*100, y = accCh*100, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = T) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v2[2:3]) + 
    scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("delta SMF (1 - CA) (", CL2," - ", CL1, ") [%]"),
         x = paste0("delta 5mC (", CL1," - ", CL2, ") [%]"),
         caption = paste("R antagonist =", round(cor_val$`accCh_antagonist`[8],3), "| n = ", numbers$n)) +
    theme_EK() +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

FILENAME = paste("AccChange", CL2, "lm", "antagonist", sep="_")
ADD = "_v1_"
# ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".png"), height = 4, width = 6.5)
# ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".pdf"), height = 4, width = 6.5)
```
```{r}
# Only agonist
numbers <- data_tbl2 %>% 
  filter(state2 == "agonist") %>% 
  count(state2)
  
p_AccCh <- data_tbl2 %>% 
    filter(state2 == "agonist") %>% 
    ggplot(aes(x = NWCGW_change*100, y = accCh*100, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = T) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v3[2:5]) + 
    scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("delta SMF (1 - CA) (", CL2," - ", CL1, ") [%]"),
         x = paste0("delta 5mC (", CL1," - ", CL2, ") [%]"),
         caption = paste("R agonist =", round(cor_val$`accCh_agonist`[9],3), "| n = ", numbers$n)) +
    theme_EK() +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

FILENAME = paste("AccChange", CL2, "lm", "agonist", sep="_")
ADD = "_v1_"
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".png"), height = 4, width = 6.5)
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".pdf"), height = 4, width = 6.5)
```

```{r}
# Only neutral
numbers <- data_tbl2 %>% 
  filter(state2 == "neutral+") %>% 
  count(state2)
  
p_AccCh <- data_tbl2 %>% 
    filter(state2 == "neutral+") %>% 
    ggplot(aes(x = NWCGW_change*100, y = accCh*100, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = T) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v3[2:3]) + 
    scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("delta SMF (1 - CA) (", CL2," - ", CL1, ") [%]"),
         x = paste0("delta 5mC (", CL1," - ", CL2, ") [%]"),
         caption = paste("R neutral =", round(cor_val$`accCh_neutral+`[7],3), "| n = ", numbers$n)) +
    theme_EK() +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

FILENAME = paste("AccChange", CL2, "lm", "neutral", sep="_")
ADD = "_v1_"
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".png"), height = 4, width = 6.5)
ggsave(plot = p_AccCh, paste0(WD, OUTDIR, FILENAME, ADD, DATE, ".pdf"), height = 4, width = 6.5)
```

\

**log2(FC) accessibility**
```{r}
p_AccCh <- data_tbl2 %>% 
    ggplot(aes(x = NWCGW_change*100, y = log2FC, color = state2)) +
    geom_hline(color = "black", yintercept = 0, linetype = "dotted") +
    geom_point(alpha = 0.8, size = 2, show.legend = F) +
    geom_smooth(color = "grey20", method = "lm") +
    scale_color_manual(values = COLORS_STATE_v3[2:5]) + 
    # scale_y_continuous(limits = c(-40, 40)) +
    labs(subtitle = paste("accessibility change in", CL1, "vs", CL2),
         y = paste0("log2 (fold change in accessibility) (", CL2," - ", CL1, ")"),
         x = paste0("delta 5mC (", CL2," - ", CL1, ") [%]"),
         caption = paste("R neutral =", round(cor_val$`log2FC_neutral+`[7],3),
                         "\nR antagonist =", round(cor_val$log2FC_antagonist[8],3),
                         "\nR agonist =", round(cor_val$log2FC_agonist[9],3))) +
    theme_EK() +
    facet_wrap(~state2, ncol = 2) 
p_AccCh

```

\

##### Linear model
```{r}
df_A <- data_f2 %>% 
  mutate(NWCGW_change = TETTKO_NO_wmean_me_bin - ES_NO_wmean_me_bin,
         DGCHN_change = -(TETTKO_NO_wmean_acc_bin - ES_NO_wmean_acc_bin)) %>%
  filter(state2 == "antagonist")

TETTKO_A.lm <- lm(DGCHN_change ~ NWCGW_change, data = df_A)
summary(TETTKO_A.lm)

df_N <- data_f2 %>% 
  mutate(NWCGW_change = TETTKO_NO_wmean_me_bin - ES_NO_wmean_me_bin,
         DGCHN_change = -(TETTKO_NO_wmean_acc_bin - ES_NO_wmean_acc_bin)) %>%
  filter(state2 == "neutral+")

TETTKO_N.lm <- lm(DGCHN_change ~ NWCGW_change, data = df_N)
summary(TETTKO_N.lm)
```
