---
title: "TF centric analysis - final data table - ES, TKO_DE, TET"
author: "Elisa Kreibich"
date: "19/09/2022"
output: 
  html_notebook:
    code_folding: hide
editor_options:
  chunk_output_type: inline
---
**Disclaimer about nomenclature**  
In the manuscript, we call the different states "sites with xyz 5mC-CA-association". Here, we call them in their original nomenclature:  

* antagonist = negative 5mC-CA-association  
* neutral = no 5mC-CA-association  
* agonist = positive 5mC-CA-association  


# Introduction
### Aim
This script calculates coverage and methylation states for the CpG (NWCGW) and all GpC (DGCHN) in the three bins at a TFBS. Then it combines this with the additional information (chromHMM, Cochran-Mantel-Haenszel test results) and calls the states to identify antagonist and neutral sites. 
In addition, coverage and methylation data of TKO (DNMT TKO / TET TKO) are calculated and added.  
The words CpG/NWCGW and GpC/DGCHN are used interchangeably in this script.  

\

### Additional information on this version:  

* 10x coverage cutoff for SM analysis  
* 5x coverage cutoff per fraction (at least 5 reads in closed and bound fraction each)
* 30 bp collection window around TFBS
* bin must be covered by at least 2 replicates  

* FINAL data table contains:  
  * **TFBS** - TFBS number
  * **motif** - motif name  
  * **motif_ID** - motif ID from motif clustering  
  * **representative.motif** - representative motif name of motif cluster  
  * **state** - neutral/antagonist (neutral == everything that is NOT antagonist)  
  * **state2**  - neutral/neutral+/antagonist (neutral+ == only TFBS around 0 of -log(COR) == conservative neutrals)  
   * **COR**, **pval** - statistics from Cochran-Mantel-Haenszel (CMH) test (COR = common odds ratio) 
  * **fraction** - closed/bount/accessible  
  * **mean_cov_TFBS** - replicates mean of TFBS coverage  
  * **wmean_me_TFBS** - replicates weighted mean of TFBS average 5mC  
  * **wmean_me_fraction** - replicates weighted mean of fractions average 5mC  
  * **mean_cov_fraction** - replicates mean of fraction coverage  
  * **ratio_mean_cov_fraction** - ratio of fractions from replicates mean  
  * **seqnames**, **start**, **end**, **width**, **strand** - genomic location of TFBS  

\

# Data Analysis
## Set environment
```{r}
WD = '/g/krebs/kreibich/Kreibich_2023_5mC_at_enhancers/'
```

### Load libraries
```{r load libraries, include=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(GenomicRanges))

#Load arguments
source(paste0(WD, 'scripts/utilities/Input_arguments_TFBS.R'))
```

### Set arguments
```{r arguments}
INDIR           <- 'data_results/single_molecule_call'
INDIR_rds       <- 'data/'
OUTDIR          <- 'data_results/'
DATE            <- Sys.Date()

SAMPLENAMES = c("ES_NO", "TKO_DE", "TETTKO_NO")
REPS = list(c("R1", "R2", "R5a6"), c("R1a2", "R3a4", "R5a6"),  c("R1", "R2"))

SAMPLENAMES_1 <- c("SMF_MM_ES_NO_R1", "SMF_MM_ES_NO_R2", "SMF_MM_ES_NO_R5a6")
SAMPLENAMES_2 <- c("SMF_MM_TKO_DE_R1a2", "SMF_MM_TKO_DE_R3a4", "SMF_MM_TKO_DE_R5a6")
SAMPLENAMES_3 <- c("SMF_MM_TETTKO_NO_R1", "SMF_MM_TETTKO_NO_R2")

SAMPLEREPS_1 <- str_extract(SAMPLENAMES_1[-1], pattern = "R.{1,3}")

FINAL_TBL_NAME <- paste('Final_data_tibble_TF', DATE, 
                         paste(SAMPLENAMES, collapse = "_"),
                         paste0("cOTFBS", cO_TFBS), 
                         sep = '_')

```

### Load in data
#### TFBS & uniqueness data
```{r load data - TFBS, message=FALSE}
if(!exists("MotDBf")){
  MotDBf <- readRDS(paste0(WD, INDIR_rds, 'mapped_jaspar2018_ChIP_score10_inBaits_BANP_all_locus_NONunique.rds'))
  MotDBf_tbl <- as_tibble(MotDBf) %>%
    mutate(motif = str_replace_all(name, "::", "_")) %>%
    select(-name)
}
```

#### CMH (Cochran-Mantel-Haenszel) test data
```{r load data - CMH data, message=FALSE}
# CMH single molecule data
CMH_tbl <- readRDS(paste0(WD, INDIR, 'TF_CMH_matrix_', 
                               SAMPLENAMES_1[1], '_', paste(SAMPLEREPS_1, collapse = '_'), 
                               '_TF_jaspar2018_BANP_inM', WDW_SIZE, '_Co', cO_TFBS, '_pc', ADD,'.rds'))
```

#### ChromHMM data
```{r}
chromHMM <- readRDS(paste0(WD, INDIR_rds, 'chromHMM_', WDW_SIZE,'bp_TFBS_with_clusters.rds')) %>% 
  select(bin, chromHMM, chromHMMcl) %>% 
  dplyr::rename(TFBS = bin)
```

## Data transformation
### Methylation and coverage data
```{r}
input_cov_tbl_l <- lapply(seq(3), function(i) {
  # i = 1
  x1 <- readRDS(paste0(WD, INDIR, 'TF_Coverage_matrix_SMF_MM_', 
                       SAMPLENAMES[i], '_', paste(REPS[[i]], collapse = '_'), '_TF_jaspar2018_BANP_inM', WDW_SIZE, '.rds'))

  x1.2 <- x1 %>% 
    add_count(sample, bin, state, name = "state_count") %>% 
    filter(state_count == 1, value == 0) %>% 
    select(1:3) %>% 
    mutate(value = 1, n = 0)
  
  x1.3 <- x1 %>% 
    full_join(., x1.2)
  
  #Count reads per fraction per state (me/unme) 
  x2 <- x1.3 %>% 
    dplyr::rename(TFBS = bin,
                  fraction = state,
                  cov_value = n)
  
  #Count reads per fraction
  x3 <- x2 %>% 
    add_count(sample, TFBS, fraction, wt = cov_value, name = "cov_fraction")
  
  #Count reads per TFBS
  x4 <- x3 %>% 
    add_count(sample, TFBS, wt = cov_value, name = "cov_TFBS")
  
  #Calculate methylation per fraction
  x5 <- x4 %>% 
    filter(value == 1) %>% 
    mutate(me_fraction = cov_value/cov_fraction) %>% 
    group_by(sample, TFBS) %>% 
    mutate(me_TFBS = sum(cov_value)/cov_TFBS) %>% 
    ungroup() %>% 
    select(-c(value, cov_value)) %>% 
    distinct(sample, TFBS, fraction, .keep_all = TRUE)
  
  #Keep only bin that are covered by at least two replicates
  x5.1 <- x5 %>% 
    distinct(sample, TFBS) %>% 
    count(TFBS) %>% 
    filter(n > 1) %>% 
    pull(TFBS) 
  
  x6 <- x5 %>% 
      filter(TFBS %in% x5.1)
  
  #Filter by bin coverage cutoff of 10 for each replicate & Separate sample & replicate
   x6.1 <- x6 %>% 
    filter(cov_TFBS >= cO_TFBS) %>% 
    mutate(replicate = str_extract(sample, "R.{1,3}"),
           sample = str_extract(sample, "(?<=SMF_MM_)\\w{1,}(?=_R.{1,})"))

  #REPLICATE INFO: Calculate weighted mean of methylation & mean of coverage
    x6.2 <- x6.1 %>% 
      distinct(TFBS, replicate, .keep_all = TRUE) %>% 
      group_by(TFBS) %>% 
      mutate(mean_cov_TFBS = mean(cov_TFBS, na.rm = T),
             wmean_me_TFBS = weighted.mean(me_TFBS, cov_TFBS, na.rm = T)) %>% 
      ungroup() %>% 
      select(contains("TFBS"), contains("ratio"))
    
    x7 <- x6.1 %>% 
      left_join(., x6.2) %>% 
      group_by(TFBS, fraction) %>% 
      mutate(wmean_me_fraction = weighted.mean(me_fraction, cov_fraction, na.rm = T),
             mean_cov_fraction = mean(cov_fraction, na.rm = T),
             ratio_mean_cov_fraction = mean_cov_fraction/mean_cov_TFBS) %>% 
      ungroup() %>% 
      select(-c(starts_with("cov"), starts_with("me_"), replicate)) %>% 
      distinct(TFBS, fraction, .keep_all = T) 
  
  #Add sample name to column names
  x8 <- x7 %>% 
    pivot_wider(names_from = "sample", values_from = contains("mean"), names_glue = "{sample}_{.value}")

})

x10 <- input_cov_tbl_l[[1]] %>%
  left_join(., input_cov_tbl_l[[2]]) %>%
  left_join(., input_cov_tbl_l[[3]])

x10$TFBS %>% unique() %>% length() #148,993
```

  
#### Filter for coverage and bound
```{r}
TFBS_cO_fraction_not_met <- x10 %>% 
    filter(fraction %in% c("bound", "closed")) %>% 
    filter(ES_NO_mean_cov_fraction < cO_fraction) %>% 
    pull(TFBS) %>% 
    unique() #41,636

TFBS_bound <- x10 %>% 
  pivot_wider(names_from = "fraction", values_from = contains("_fraction"), names_sep = "_") %>% 
  filter(ES_NO_ratio_mean_cov_fraction_bound >= cO_BOUND) %>% 
  pull(TFBS) %>% 
  unique() #11,247

x11 <- x10 %>% 
  # Filter for overall coverage  
  filter(ES_NO_mean_cov_TFBS >= cO_TFBS) %>% 
  # Filter for fraction coverage  (bound | closed)
  filter(!(TFBS %in% TFBS_cO_fraction_not_met)) %>% 
  # Filter for 5% bound
  filter(TFBS %in% TFBS_bound)

x11$TFBS %>% unique() %>% length() #5332
```

### Add other data
```{r}
x12 <- x11 %>% 
  #Add CMH tibble
  left_join(., CMH_tbl) %>% 
  #Add ChromHMM
  left_join(., chromHMM) %>% 
  #Add TFBS & motif cluster info
  left_join(., MotDBf_tbl) %>% 
  filter(!is.na(seqnames))
  
x12$TFBS %>% unique() %>% length() #5,249

```


### Annotate state
```{r}
# Get intermediately methylated data and recreate states
COI <- names(x12)[grepl(paste0(SAMPLENAMES[1], "_wmean_me_TFBS"), names(x12))]

data_iMe <- x12 %>% 
  filter(dplyr::between(get(COI), AVME.i.1, AVME.i.2)) %>% 
  mutate(state2 = case_when(
                      COR <= COR.i.antago & pval < PVAL.i ~ "antagonist",
                      COR >= COR.i.ago & pval < PVAL.i ~ "agonist",
                      dplyr::between(COR, COR.i.neut.1, COR.i.neut.2) ~ "neutral+",
                      TRUE ~ "neutral")) %>% 
  mutate(state = case_when(
                      state2 == "neutral+" ~ "neutral", 
                      TRUE ~ state2)) %>% 
  mutate(state_crude = case_when(
                      COR <= COR.i.antago ~ paste("COR <", COR.i.antago),
                      COR > COR.i.antago ~ paste("COR >", COR.i.antago)))

# Get NOT intermediately methylated data and create states
data_NOTiMe <- x12 %>% 
  filter(get(COI) < AVME.i.1 | get(COI) > AVME.i.2) %>% 
  mutate(state2 = case_when(
                    ES_NO_wmean_me_TFBS < AVME.i.1 ~ "unmethylated",
                    ES_NO_wmean_me_TFBS > AVME.i.2 ~ "fully methylated"),
         state = state2) %>% 
  mutate(state2 = case_when(
                    get(COI) < AVME.i.1 ~ "unmethylated",
                    get(COI) > AVME.i.2 ~ "fully methylated"),
         state = state2,
         state_crude = state2)
  
#Combine 
datatbl <- rbind(data_iMe, data_NOTiMe) %>% 
  mutate(state = factor(state, levels = LEVELS_state),
         state2 = factor(state2, levels = LEVELS_state2),
         state_crude = factor(state_crude, levels = LEVELS_state_crude)) %>% 
  mutate(fraction = factor(fraction, levels = c("bound", "closed", "accessible"))) %>% 
  mutate(across(contains("mean"), round, 3))

datatbl$TFBS %>% unique() %>% length() #5249
datatbl %>% summary()
```

### Save final table
```{r}
#Save final data table
saveRDS(datatbl, paste0(WD, OUTDIR, FINAL_TBL_NAME, '.rds'))

#As GRange object
suppressPackageStartupMessages(library(plyranges))
datatbl_gr <- as_granges(datatbl)
saveRDS(datatbl_gr, paste0(WD, OUTDIR, FINAL_TBL_NAME, '_gr.rds'))
```
